---
title: "Instructions for setting up RStudio running in a Docker container on Jetstream or Digital Ocean, including multiple user logins"
author: "Dana Miller"
date: "February 24, 2020"
output:
  github_document:
    toc: true
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


*With thanks to the [Rocker Project](https://www.rocker-project.org/) for R on Docker, Indiana University and the National Science Foundation for cloud compute resources through [Jetstream](https://jetstream-cloud.org/), [DIBSCI](http://ivory.idyll.org/dibsi/2018/WHO.html) at UC Davis for documentation, and Carl Boettiger at UC Berkeley for reproducible research inspiration and an initial version of these instructions*


### Background - two cloud compute providers referenced in this guide
- [Jetstream](https://jetstream-cloud.org/) is an National Science Foundation supported cloud compute resource that runs on the open-source software Atmosphere and OpenStack. Compute credit allocations are available by applying through XCSEDE. If you can contact campus research computing staff, they may be able to provide advice and example applications (eg at UC Berkeley this is the [Research Computing](https://research-it.berkeley.edu/programs/berkeley-research-computing) group)  

-[Digital Ocean](https://www.digitalocean.com/) is a commercial cloud compute resource (similar to AWS or Azure) offering similar instances to those available on Jetstream. Compute credits are billed hourly, depending on the size of the resource. 

Running RStudio in Docker and configuring multiple user logins works on both of these platforms. Most of the instructions below apply to both platforms, anything that is Jetstream or Digital Ocean-specific will be noted. 

### Set up a new instance for the first time

#### On Jetstream
If you already have a Jetstream login and allocation, to run R Studio on Jetstream:

   - Log in and go to 'Create a new project'
   - Enter project name and description
   - Select the size and *image* you would like to use  (images have various software preinstalled): select [this](https://use.jetstream-cloud.org/application/images/107) Ubuntu image that includes Docker and GUI support
   - Once the new instance for your project is active (green dot), click the "Open web shell" link on the right hand side
   
#### On Digital Ocean

- After creating a Digital Ocean account, create a new instance using [this](https://marketplace.digitalocean.com/apps/docker) Ubuntu + Docker image (click 'create Droplet')
- Recommend setting up SSH authentication
- Once the droplet is set up, connect to the droplet's shell interface by opening a terminal and connecting via ssh with 
```
ssh root@[IP address of your droplet here]
```

### Start a Docker container with R and RStudio 

- In the web shell, first confirm you are able to run docker commands: 
    
    ```{bash, eval = FALSE}
    docker run hello-world
    ```
       
   If configured correctly, this will say `Hello from Docker!...`
  
   If you get an error about permission denied and docker daemon, try adding yourself to root privileges with `sudo usermod -aG docker example_username`, where `example_username` is your username
        
   - Start a new Docker container that has R, R Studio, and the `tidyverse` preinstalled (you could also install these and all the required dependencies without Docker but it takes a long time and is potentially error-prone)
    
```{bash, eval = F}
    docker run -d -p 8787:8787 --rm -e PASSWORD=example_password rocker/tidyverse
```

or, if in addition you also want TeX installed and tools for publishing, use the larger `verse` image:

```{bash, eval = F}
    docker run -d -p 8787:8787 --rm -e PASSWORD=example_password rocker/verse
```


*What do all those flags mean?*
    
 - "-d" for *detach* means the container will run 'in the background' , and return a command prompt (without this flag your container will be running but you won't be able to run anything else at the command prompt until the container is stopped)
 - "-p" for *port* is a port and also the suffix for the URL your RStudio session will be accessible at
 - "-e" for *environment* is passing the `PASSWORD` variable to the created computational environment
 - "--rm" for *remove* , which will automatically delete this container after it exits (note it does not delete the underlying image the container was based on)
 - `rocker/tidyverse` - this specifies the exact Docker image (already created by the Rocker project) to use. Using the same image, you can set up the exact same computational environment many different times, or many different people can all use it to set up the same environment
                - More details about this image at [https://www.rocker-project.org/images/](https://www.rocker-project.org/images/)
    
  - To find the URL you can use to log into the new R Studio session, run 
    
    ```{bash, eval = FALSE}
    echo My RStudio Web server is running at: http://$(hostname):8787/
    ```
    
Note the URL above is *http* not *https*, the latter is also possible to set up but requires a custom domain name

Note to copy and paste text from the Jetstream web shell, you have you highlight it and then press Ctrl + Alt + Shift
    
  - Paste the URL above into your web browser, and to login:
        
     - The default username is "rstudio"
     - The password is whatever you chose in the `docker run...` command above for `example_password`
        
    Congrats, you are now running RStudio on Jetstream!
    
    The `rstudio` user has permission to install packages. 
    
   
   Optional UNIX configuration notes: 
   
   - `whoami` in at the command line is *not* the same as the user for logging in to the R Studio browser prompt 
   - Specifying a username with `docker run -u...` also doesn't work for the R Studio login prompt

### Configure git with `git config`

If you will be using `git` to incrementally save your work on the instance and then push the saved changes to a remote repository like GitHub, each user who separtely logs into the RStudio instance will need to run the two git configuration commands below in the RStudio Terminal pane. 

```
 git config --global user.name "Your Name"
 git config --global user.email your_username@users.noreply.github.com
```

Note that for the user email, you can either use your own email, or use the 'noreply' email generated for you by GitHub to avoid exposing your personal email address in your commit history. See [here](https://help.github.com/en/github/setting-up-and-managing-your-github-user-account/setting-your-commit-email-address) for details. 

    
### Optional next step - adding multiple users and configuring their permissions

There are two ways to do this: 

1. Add users to your existing instance, or
2. Launch a separate R Studio instance bound to a different port


#### 1. Add more users to an existing instance

- With your Jetstream instance active, launch the web shell
- See the ids of your running Docker container with RStudio by running `docker ps`. You will need to write down or remember the first 5 digits of your CONTAINER ID (a long string of numbers and letters)

- To run commands inside that container, run

```{bash, eval = FALSE}

    docker exec -ti CONTAINER_ID bash
```

Where CONTAINER_ID is eg "904b2",  the first 5 digits. Notice your prompt will change now that you are inside the container.

- Now you can use regular UNIX commands to create a new user (this is not Docker-specific), where `example_user` is the user name you are creating (the name can be anything you choose). 

```{bash, eval = F}
 adduser example_user
```

Enter a password, press 'Enter' to leave the other questions blank (eg Name, Work Phone...), they are not required, and press Y for "Yes" 

- At this point, someone could go to the same URL you set up above, and login with this new username and password *but* they wouldn't be able to install any additional R packages (no root access)

- To give the new user the right to download packages, add them to the `staff` group. In the command below, `-aG` means 'append to group'

```{bash, eval = F}
 usermod -aG staff example_user
```

 To see a list of all the users (this includes some other system info) as well

```{bash, eval = F}
cat /etc/passwd
```

To change a password (note `passwd` below is not a typo)

```{bash, eval = F}
sudo passwd example_username
```

When you are done configuring the user info, type 
```{bash, eval = F}
exit
```
to return to the command line outside of that container. 

And the login URL for the new users same as above, at: 

```{bash, eval = FALSE}
echo My RStudio Web server is running at: http://$(hostname):8787/
```


#### 2. Launch a separate R Studio instance bound to a different port

Alternatively, start an additional Docker container on the same Jetstream instance with a different port number

 ```{bash eval = F}
 docker run -d -p 8888:8787 -rm -e PASSWORD=newpassword rocker/tidyverse 
 ```

The default username will still be "rstudio", and the password will be whatever you just specified, but the login URL will be different (note the port number is changed above and accordingly also changed below) 
    
```{bash, eval = FALSE}
echo My RStudio Web server is running at: http://$(hostname):8888/
```

### Optional next steps - installing `curl`, `pip3` and `aws cli` in the terminal after RStudio is running

- [`curl`](https://curl.haxx.se/) is a common command line utility that you will probably need if you are downloading data or files from URLs
- `pip3` is the package manager for the `python` programming language. In this case, it's required to install `aws cli`
- `aws cli` is a command line interface for Amazon Web Services

The following steps were tested on a fresh `verse` Docker image.

In the 'Terminal' pane of RStudio : 

1. Install curl, eg following instructions [here](https://www.luminanetworks.com/docs-lsc-610/Topics/SDN_Controller_Software_Installation_Guide/Appendix/Installing_cURL_for_Ubuntu_1.html)

```
sudo apt-get update
```

```
sudo apt-get install curl
```

```
curl --version
```

2. Install pip - use python 3

- See "Install pip" instructions [here](https://docs.aws.amazon.com/cli/latest/userguide/install-linux.html#install-linux-pip)
    - `verse` image has python three, so use the python3 variants, eg `python3 get-pip.py --user`, not `python get-pip.py --user`
    - For the step about editing your bash profile, it will probably be called `.profile`. 
    - You can use `vim` to edit the file (type `vim .profile` to open the file: 
    
        - `i` to go into "insert" (editing) mode
        - add `export PATH=~/.local/bin:$PATH` to the file below anything in it already (you can start a line with `#` as a comment to explain why it's there)
        - press `:` to exit editing mode and `wq` to save (write) and quit to go back to the terminal
    - Per the AWS docs, make sure you source the file after editing it! ` source ~/.bash_profile`
        
3. Install `aws cli`

Continue to follow instructions [here](https://docs.aws.amazon.com/cli/latest/userguide/install-linux.html#install-linux-awscli)
    
### Optional next steps - installing most R packages
#### Packages on CRAN (most R packages you'll need)
- The `verse` image has the `tidyverse` packages preinstalled, which you can see with 

```{r, eval = F}
library(tidyverse)
```


- Most additional R packages can be installed as usual from CRAN in the R console, eg with 

```{r, eval = F}
install.packages('skimr')  # skimr is the name of the package
                           # package names must be in quotes
```

- Packages we commonly install that are not pre-installed in the `verse` Docker image include: `skimr`, `cowplot`, `here`, and `janitor`

#### Packages on GitHub

- If you encounter a package that is not available on CRAN, eg [`aws.s3`](https://github.com/cloudyr/aws.s3) from the [`cloudyr`](https://cloudyr.github.io/) project (which as of this writing was [removed from CRAN](https://twitter.com/thosjleeper/status/1216426950413234177) in January 2019 for lack of an ongoing maintainer ), 1) An archive may be available from CRAN (check the package's old CRAN page), 2) It is probably available on GitHub. 

- To install `aws.s3` from GitHub, run (and can see more detailed instructions [here](https://github.com/cloudyr/aws.s3))

```
install.packages("aws.s3", repos = c("cloudyr" = "http://cloudyr.github.io/drat"))
```

### Optional next steps - installing `sparklyr` and `spark` (required reinstalling Java 😱, but you can do it 💪 )

First, install `sparklyr`, and use it to install spark
```{r, eval = F}
install.packages(`sparklyr`)

sparklyr::spark_install()
```

If you try and create a spark connection and get an error about installing Java 8, see below

#### Context - what is the problem with Java 8 vs Java 11?

- As documented by other sparklyr users [here](https://github.com/uc-cfss/Discussion/issues/71), and the ['Mastering Spark in R'](https://therinspark.com) book [here](https://therinspark.com/starting.html), `sparklyr` requires the Java Runtime Environment, version Java 8. 
- However, the Docker container from the Rocker project that we just set up RStudio in comes with a different version, Java 11. Specifically, the container is using the Debian version of the Linux, and (as of this writing) the current version of Debian, Debian 10, comes with OpenJDK version 11, which is an open-source version of the Java Runtime Environment. 
- Note this does mean that even though your Jetstream or Digital Ocean instance is running the Ubuntu version of Linux, the RStudio session running *inside the Docker container* is running in the Debian version of Linux. Installing software on Debian and Ubuntu can be different via the command line, so make sure to specify the version of Linux when you are searching for installation help. 

You can check your java version in the terminal with:

```
java -version
```

The steps below follow the steps suggested in the ['Java errors with sparklyr?'](https://github.com/uc-cfss/Discussion/issues/71) issue linked above

#### Removing OpenJDK Java 11

- [Here](https://java.com/en/download/help/linux_uninstall.xml) are the official Java instructions on uninstalling Java on Linux
- The Docker container appeared not to have an RPM install, after searching for the java install location  with `which java` I removed it with 

```
sudo rm -r /usr/lib/jvm/java-11-openjdk-amd64
```

and confirmed java was uninstalled with 


```
java -version
```

#### Installing OpenJDK Java 8 on Debian 10

- Unfortunately, this will not work, since [apparently](https://serverfault.com/questions/974992/why-isnt-there-a-openjdk-8-jdk-package-on-debian-anymore) it's not available by default for Debian 10
```
# Seems like it might work, but will not
apt-get install openjdk-8-jdk
```


- Fortunately, there is an alternative. I followed instructions from AdoptOpenDK  [here](https://adoptopenjdk.net/installation.html#) for installing OpenJDK on Debian here (specifically the "DEB installation on Debian or Ubuntu")

- Per the messages from following those instructions, first I needed to install `gnupg2`, which is "a complete and free implementation of the OpenPGP standard" ([link](https://gnupg.org/) to project), which is needed for the step where you get the GPG key below

```
sudo apt-get install gnupg2
```

- Then could follow these steps from the instructions above

```
# Get AdoptOpenJDK GPG key 
wget -qO - https://adoptopenjdk.jfrog.io/adoptopenjdk/api/gpg/key/public | sudo apt-key add -

#  Import AdoptOpenJDK DEB repository
sudo add-apt-repository --yes https://adoptopenjdk.jfrog.io/adoptopenjdk/deb/

#refresh package list
apt-get update

#install
apt-get install adoptopenjdk-8-hotspot

```

Check your java version again:

```
java -version
```

Now it should say `openjdk version 1.8`, eg 

```
openjdk version "1.8.0_242"
OpenJDK Runtime Environment (AdoptOpenJDK)(build 1.8.0_242-b08)
OpenJDK 64-Bit Server VM (AdoptOpenJDK)(build 25.242-b08, mixed mode)
```

Now 1) Restart your R session
    2) Try re-running `config <- spark_config()` - hopefully this resolves the issue for you too 🎉


### More references

- DIBSCI docs on [setting up a Jetstream instance](https://angus.readthedocs.io/en/2018/jetstream/boot.html) and   [installing and running RStudio on Jetstream](https://angus.readthedocs.io/en/2018/visualizing-blast-scores-with-RStudio.html#installing-and-running-rstudio-on-jetstream) - note some of the instructions are specific to their workshop
- See more details from the Rocker Project on managing user authentication [here](https://www.rocker-project.org/use/managing_users/)
- [Docker docs](https://docs.docker.com/)